### 2016-07-08 appendix.tex
Document structure changes and new appendix changes
Persistent directories
New appendix and including more description about the persistent directory setup

\label{sec:testbed_http}

\section{Old version of losetup}

\begin{lstlisting}[]
# export DEV="${losetup -f}"
/dev/loop0
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

\begin{lstlisting}[]
# mknod -m 660 ${DEV}p1 b 7 1
# mknod -m 660 ${DEV}p2 b 7 2
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

\begin{lstlisting}[]
# fdisk -lu ${IMAGE}.img
WHAT WILL BE THE OUTPUT?
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

I THINK SIZELMIT IS NOT REQUIRED
\begin{lstlisting}[]
# losetup -o $((8192*512)) --sizelimit $((137215*512)) ${DEV}p1 ${IMAGE}.img
# losetup -o $((137216*512)) --sizelimit $((4806655*512)) ${DEV}p2 ${IMAGE}.img
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

The output from \texttt{lsblk} will look a bit different

>>>>

\section{Advanced Setup: HTTP Image Fetching}
\label{sec:http}

The HTTP description

\section{Alternative Commands for Outdated Packages}
This section describes alternative commands in case the testbed administrator
is not able to update old packages in its Linux distribution for performing
the commands, particularly the ones regarding the \texttt{util-linux} package.

\subsection{Losetup for Loopback Devices}
\label{sec:alternative_losetup}

In case the \texttt{losetup -P} is not available given that a version
higher or equal than 2.21 for \texttt{util-linux} is neither not available,
a possible option is to manually set all the loopback devices used during the
whole procedure with \texttt{mknod}. To do this, the alternative commands are:

\begin{lstlisting}[]
$ export DEV="/dev/loop0"
# mknod ${DEV}p1 b 7 1
# mknod ${DEV}p2 b 7 2
# losetup -o $((8192*512)) --sizelimit $((137215*512)) ${DEV}p1 ${IMAGE}.img
# losetup -o $((137216*512)) --sizelimit $((4806655*512)) ${DEV}p2 ${IMAGE}.img
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

However, there will be a few differences by using the above code as an
alternative to the \texttt{losetup -P} case. First, the output from
\texttt{lsblk} will look different:

\begin{lstlisting}[]
# lsblsk
  NAME   MAJ:MIN RM   SIZE RO TYPE MOUNTPOINT
  sda      8:0    0 238.5G  0 disk
    sda1   8:1    0   512M  0 part /boot
    sda2   8:2    0   168G  0 part /
    sda3   8:3    0     2G  0 part [SWAP]
    sda4   8:4    0    68G  0 part
  loop1    7:1    0    67M  0 loop
  loop2    7:2    0   2.2G  0 loop
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Second, detaching the loop devices will also be different:

\begin{lstlisting}[]
# losetup -d ${DEV}p{1,2}
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Working with these loopback devices will be transparent for the remainding
procedure.

Also, when creating the persistent home directory partition in
Section~\ref{sec:persistent_directories}, for recreating the loopback devices
the commands are:

\begin{lstlisting}[]
$ export DEV="/dev/loop0"
# mknod ${DEV}p2 b 7 2
# mknod ${DEV}p3 b 7 3
# losetup -o $((137216*512)) --sizelimit $(( (4806655-137216+1)*512)) ${DEV}p2 ${IMAGE}.img
# losetup -o $((4806656*512)) --sizelimit $(( (6903807-4806656+1)*512)) ${DEV}p3 ${IMAGE}.img
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Finally, for detaching in this case the command is:
\begin{lstlisting}[]
# losetup -d ${DEV}p{2,3}
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

\subsection{Umount after Image Chroot}
\label{sec:umount}

In case the \texttt{umount --recursive} is not available given that a version
higher or equal than 2.22 for \texttt{util-linux} is neither not available,
a possible option is to manually umount all the partitions used during the
\texttt{chroot} environment. The directories must be unmounted in the reverse
order as when they were mounted. These commands are:

\begin{lstlisting}[]
# umount ${ROOTDIR}/dev/pts
# umount ${ROOTDIR}/dev
# umount ${ROOTDIR}/sys
# umount ${ROOTDIR}/proc
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

### 2016-07-08 overlay_fs.tex
New appendix and including more description about the persistent directory setup

For the \texttt{--recursive} to work properly, it is necessary that
the package \texttt{util-linux} version is greater than or equal to 2.22.
Otherwise, an alternative is to either update the package or follow the
procedure in Section~\ref{sec:umount} of the Appendix.

\subsection{Persistent Image Directories}
Provided the stacked filesystem is configured, now it is possible to have
directories where files are not removed upon rebooting the \ac{Raspi}s.
The following procedure creates an extra partition in the image for
the \ac{Raspi} user home directory that will be made storage persistent
using the upper filesystem. We first expand image according to the desired
home directory size, but avoiding to make the image bigger than the target
memory card size.

@@

Create home partition after the root partition. Use \texttt{fdisk} to
find the next available sector in the image. In our case, the new
partition should start at sector 4806656:
\begin{lstlisting}[]
$ fdisk -lu ${IMAGE}.img
Disk 2016-05-27-raspbian-jessie-lite.img: 3.3 GiB, 3534749696 bytes, 6903808 sectors
Units: sectors of 1 * 512 = 512 bytes
Sector size (logical/physical): 512 bytes / 512 bytes
I/O size (minimum/optimal): 512 bytes / 512 bytes
Disklabel type: dos
Disk identifier: 0x6fcf21f3

@@

Create new partition:

\begin{lstlisting}[]
$ (echo n; echo p; echo ; echo 4806656; echo ; echo p; echo w) | fdisk ${IMAGE}.img
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Create loopback device:
\begin{lstlisting}[]
# export DEV=$(sudo losetup --show -f -P ${IMAGE}.img); echo $DEV
/dev/loop0
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Format new partition:
\begin{lstlisting}[]
# mkfs.ext4 ${DEV}p3
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

The last thing to remember is to inform Raspbian to mount the home partition
we created during startup. This can be done by adding an entry in \texttt{fstab}:
\begin{lstlisting}[]
# mount ${DEV}p2 ${ROOTDIR}
# sudo sh -c "echo '/dev/mmcblk0p3 /home ext4 defaults,noatime 0 2' >> ${ROOTDIR}/etc/fstab"
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

>>>>

For the \texttt{----recursive} option to work properly, it is necessary that
the package \texttt{util-linux} version is greater than or equal to 2.22.
Otherwise, an alternative is to either update the package or follow the
procedure in Section~\ref{sec:umount} of the Appendix.

\subsection{Persistent Image Directories}
\label{sec:persistent_directories}
Provided the stacked filesystem is configured, now it is possible to have
directories where files are not removed upon rebooting the \ac{Raspi}s.
The following procedure creates an extra partition in the image for
the \ac{Raspi} user home directory that will be made storage persistent
using the upper filesystem. We first expand image according to the desired
home directory size, but avoiding to make the image bigger than the target
memory card size.

@@

We create a partition for the home directory after the root partition.
To do this, we again use \texttt{fdisk} to find the next available sector
in the image. To verify the new available space for the full image and observe
the next available sector, we run:
\begin{lstlisting}[]
$ fdisk -u sectors -l ${IMAGE}.img
Disk 2016-05-27-raspbian-jessie-lite.img: 3.3 GiB, 3534749696 bytes, 6903808 sectors
Units: sectors of 1 * 512 = 512 bytes
Sector size (logical/physical): 512 bytes / 512 bytes
I/O size (minimum/optimal): 512 bytes / 512 bytes
Disklabel type: dos
Disk identifier: 0x6fcf21f3

@@

We notice that one GB is now available to be used in the partitions. Also, we
observe the new partition should start at sector 4806656. To create it, we
show the introduced command and its output:

\begin{lstlisting}[]
$ (echo n; echo p; echo 3; echo 4806656; echo ; echo w) | fdisk ${IMAGE}.img

Command (m for help): Partition type:
   p primary (1 primary, 0 extended, 3 free)
   e extended
Partition number (1-4, default 3): Using default value 3
First sector (2048-6903807, default 2048): Last sector, +sectors or
+size{K,M,G} (4806656-6903807, default 6903807): Using default value 6903807

Command (m for help): The partition table has been altered!

Syncing disks.
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

We create again a loopback device and format the new partition, as follows:
\begin{lstlisting}[]
# export DEV=$(sudo losetup --show -f -P ${IMAGE}.img); echo $DEV
/dev/loop0
# mkfs.ext4 ${DEV}p3
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Again, if for some reason, the \texttt{-P} option is not available for
\texttt{losetup}, we provide an alternative command line in
Section~\ref{sec:alternative_losetup}. Finally, if the previous filesystem
formatting was successful, the filesystem is now available for use. Thus,
we need to inform the Raspbian root filesystem to mount the home partition
that we have just created. We will notify to the Raspbian \ac{OS} during
the startup of the \ac{Raspi}s. This can be done by adding an entry in
\texttt{fstab} as follows:

\begin{lstlisting}[]
# mount ${DEV}p2 ${ROOTDIR}
# sudo sh -c "echo '/dev/mmcblk0p3 /home ext4 defaults,noatime 0 2' >> ${ROOTDIR}/etc/fstab"
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}
